MSstatsPTM statistical relative quantification of post-translational modifications in global proteomics experiments \textbf{Supplementary Information}

### 2022-07-15 supplementary/ptm_sm.tex
supp text changes

The proposed statistical approach was evaluated and compared to the $t$-test and Limma methods using computer simulations. Specifically, their statistical properties with respect to protein-level adjustment and real world experimental conditions were evaluated.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Computer simulation}
\label{sec:comp_sim}
Differential levels of modified peptides may be due to differential modifications and/or changes in protein abundance. Approaches without considering protein-level adjustment lose track of an important aspect in interpreting observed changes in PTM abundance, which may result in misleading conclusions. To highlight the necessity of the adjustment, we compared six different approaches as follows: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) $t$-test (with adjustment), d) $t$-test (no adjustment), e) Limma (with adjustment), and f) Limma (no adjustment).

In experiments of complex designs, multiple inter-related conditions are often compared together. The proposed approach and Limma leverage measurements in all conditions for the inference of the underlying abundance, whereas $t$-test uses measurements from the two conditions being compared. To highlight this distinction, multiple conditions of data were generated. While multiple conditions were generated, all comparisons were still made with the same fold change between conditions ($0.75$). This ensured that all peptides were differentially abundant by the same amount and the only difference in calculations was using more conditions in the modeling.

@@

\begin{itemize}
\item Mean of log-intensity: $25$
\item Standard deviations of log-intensities for modified and unmodified peptides: $0.2$, $0.3$
\item Difference in PTM abundance between conditions: $0$, $0.75$, $1.5$, $2.25$
\item Difference in protein abundance between conditions: $0$, $0.75$, $1.5$, $2.25$
\item Number of replicates: $2$, $3$, $5$, $10$
\item Number of conditions: $2$, $3$, $4$
\item Number of realizations: $1000$
\item Number of features per PTM: $10$
\item Number of features per unmodified protein: $10$
\item Missing data: no missing value
\end{itemize}

@@

\begin{itemize}
\item Mean of log-intensity: $25$
\item Standard deviations of log-intensities for modified and unmodified peptides: $0.2$, $0.3$
\item Difference in PTM abundance between conditions: $0$, $0.75$, $1.5$, $2.25$
\item Difference in protein abundance between conditions: $0$, $0.75$, $1.5$, $2.25$
\item Number of replicates: $2$, $3$, $5$, $10$
\item Number of conditions: $2$, $3$, $4$
\item Number of realizations: $1000$
\item Number of features per PTM: $2$
\item Number of features per unmodified protein: $10$
\item Missing data: 20\% of the observations for PTMs and Proteins were masked with NA at random
\end{itemize}

@@

\begin{figure}[h!]
\centering
 \begin{subfigure}{\textwidth}
  \centering
	\includegraphics[width=.8\textwidth]{sim_new/sim1_Recall}\\
	\caption{All methods with adjustment were compared by comparing recall in simulation 1. Limma performed the strongest here when the number of replicates were low. At higher replicates the performance of the proposed methods and Limma were comparable. $t$-test clearly performed worse across all methods.}
 \end{subfigure}
 \begin{subfigure}{\textwidth}
  \centering
	\includegraphics[width=.8\textwidth]{sim_new/sim1_accuracy}
	\caption{The overall accuracy plot mimicked the observations in the recall plot. Limma performed strongest when replicates were low, but was comparable to the proposed method with more replicates.}
 \end{subfigure}
\caption{Recall and Accuracy results of Simulation 1.}
\label{fig:sim1_recall}
\end{figure}

\begin{figure}[h!]
\centering
	\includegraphics[width=.8\textwidth]{sim_new/sim3_Recall}
	\caption{}
\caption{Recall results of Simulation 2. The advantage of using the proposed approach were apparent when looking at simulation 2, which included limited observations and the presence of missing values. In the case of recall the proposed method performed stronger than Limma and t-test in nearly every model. Even at lower replicates the proposed method still outperformed Limma. Again the lowest performing method was $t$-test.}
\label{fig:sim2_recall}
\end{figure}

\begin{figure}[h!]
\centering
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.8\textwidth]{sim_new/sim1_FC_boxplot}
	\caption{In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of .75. The distributions around the median were also consistent across all methods.}
 \end{subfigure}
 \begin{subfigure}{\textwidth}
  \centering
	\includegraphics[width=.8\textwidth]{sim_new/sim3_FC_boxplot}
	\caption{In simulation 2 all methods correctly estimated the fold change with a median log change of .75. The proposed method in this simulation had a visibly tighter distribution around the median. Both Limma and $t$-test showed a wider range around the fold change. The inner quartile range of the Proposed method was on average 10.4\% smaller than t-test and 21.8\% smaller than Limma.}
 \end{subfigure}
\caption{Fold change distribution comparison between Simulations 1 and 2.}
\label{fig:fc_boxplot}
\end{figure}

@@

Again we consider three different methods and assess their performance: the method proposed in this paper, Limma, and two sample $t$-test. All methods are analyzed after adjusting for changes in overall protein level. The results are summarized from Figure \ref{fig:spike_volcano_msstats} to Figure \ref{fig:spike_stat}, including volcano plots, model summary statistics, and fold change analysis.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=1\textwidth]{sim_new/spike_in_anova_volcano}
\caption{Using the two sample $t$-test, none of the comparisons either before or after adjustment show any significant peptides. With that being said, the fold change of the spike-in peptides was much closer to expectation after adjusting for global protein abundance. \label{fig:spike_volcano_ttest}}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=.85\textwidth]{sim_new/spike_in_statistic_comparison}
\caption{Comparing the summary statistics between methods, the proposed method performed the strongest. In terms of accuracy and specificity the three methods were close, with Limma and $t$-test showing slightly higher values. Accuracy and specificity were dominated by the large number of true negatives (background peptides) compared to the true positives (spike-in peptides). In terms of recall, the proposed approach out performed the other two methods, showing that it correctly labeled the most spike-in peptides. \label{fig:spike_stat}}
\end{figure}

>>>>

The proposed statistical approach was evaluated and compared to the Anova and Limma methods using computer simulations. Specifically, their statistical properties with respect to protein-level adjustment and real world experimental conditions were evaluated.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Computer simulation}
\label{sec:comp_sim}
Differential levels of modified peptides may be due to differential modifications and/or changes in protein abundance. Approaches without considering protein-level adjustment lose track of an important aspect in interpreting observed changes in PTM abundance, which may result in misleading conclusions. To highlight the necessity of the adjustment, we compared six different approaches as follows: a) proposed approach, b) proposed approach without adjusting for unmodified peptides c) Anova (with adjustment), d) Anova (no adjustment), e) Limma (with adjustment), and f) Limma (no adjustment).

In experiments of complex designs, multiple inter-related conditions are often compared together. All comparisons were made with the same fold change between conditions ($1$). This ensured that all peptides were differentially abundant by the same amount and the only difference in calculations was using more conditions in the modeling.

@@

\begin{itemize}
\item Mean of log-intensity: $25$
\item Standard deviations of log-intensities for modified and unmodified peptides: $0.2$, $0.3$
\item Difference in PTM abundance between conditions: $0$, $1.$, $2.$, $3.$
\item Difference in protein abundance between conditions: $0$, $1.$, $2.$, $3.$
\item Number of replicates: $2$, $3$, $5$, $10$
\item Number of conditions: $2$, $3$, $4$
\item Number of realizations: $1000$
\item Number of features per PTM: $10$
\item Number of features per unmodified protein: $10$
\item Missing data: no missing value
\end{itemize}

@@

\begin{itemize}
\item Mean of log-intensity: $25$
\item Standard deviations of log-intensities for modified and unmodified peptides: $0.2$, $0.3$
\item Difference in PTM abundance between conditions: $0$, $1.$, $2.$, $3.$
\item Difference in protein abundance between conditions: $0$, $1.$, $2.$, $3.$
\item Number of replicates: $2$, $3$, $5$, $10$
\item Number of conditions: $2$, $3$, $4$
\item Number of realizations: $1000$
\item Number of features per PTM: $2$
\item Number of features per unmodified protein: $10$
\item Missing data: 20\% of the observations for PTMs and Proteins were masked with NA at random
\end{itemize}

@@

\begin{figure}[h!]
\centering
 \begin{subfigure}{\textwidth}
  \centering
	\includegraphics[width=.8\textwidth]{sim_new/sim1_Recall}\\
	\caption{All methods with adjustment were compared by comparing recall in simulation 1. Limma performed the strongest here when the number of replicates were low. At higher replicates the performance of the proposed methods and Limma were comparable. Anova clearly performed worse across all methods.}
 \end{subfigure}
 \begin{subfigure}{\textwidth}
  \centering
	\includegraphics[width=.8\textwidth]{sim_new/sim1_accuracy}
	\caption{The overall accuracy plot mimicked the observations in the recall plot. Limma performed strongest when replicates were low, but was comparable to the proposed method with more replicates.}
 \end{subfigure}
\caption{Recall and Accuracy results of Simulation 1.}
\label{fig:sim1_recall}
\end{figure}

\begin{figure}[h!]
\centering
	\includegraphics[width=.8\textwidth]{sim_new/sim3_Recall}
	\caption{}
\caption{Recall results of Simulation 2. The advantage of using the proposed approach were apparent when looking at simulation 2, which included limited observations and the presence of missing values. In the case of recall the proposed method performed stronger than Limma and t-test in nearly every model. Even at lower replicates the proposed method still outperformed Limma. Again the lowest performing method was Anova.}
\label{fig:sim2_recall}
\end{figure}

\begin{figure}[h!]
\centering
 \begin{subfigure}{\textwidth}
 \centering
	\includegraphics[width=.8\textwidth]{sim_new/sim1_FC_boxplot}
	\caption{In simulation 1 all considered methods correctly estimated the fold change between conditions, with a median fold change estimation of 1. The distributions around the median were also consistent across all methods.}
 \end{subfigure}
 \begin{subfigure}{\textwidth}
  \centering
	\includegraphics[width=.8\textwidth]{sim_new/sim3_FC_boxplot}
	\caption{In simulation 2 all methods correctly estimated the fold change with a median log change of 1. The proposed method in this simulation had a visibly tighter distribution around the median. Both Limma and Anova showed a wider range around the fold change.}
 \end{subfigure}
\caption{Fold change distribution comparison between Simulations 1 and 2.}
\label{fig:fc_boxplot}
\end{figure}

@@

Again we consider three different methods and assess their performance: the method proposed in this paper, Limma, and Anova. All methods are analyzed after adjusting for changes in overall protein level. The results are summarized from Figure \ref{fig:spike_volcano_msstats} to Figure \ref{fig:spike_stat}, including volcano plots, model summary statistics, and fold change analysis.

@@

\begin{figure}[h!]
\centering
\includegraphics[width=1\textwidth]{sim_new/spike_in_anova_volcano}
\caption{Using Anova the fold change of the spike-in peptides was much closer to expectation after adjusting for global protein abundance. The log FC estimation was the same as Limma, however the pvalues were different. We actually see a few more True Positives using Anova compared to Limma.\label{fig:spike_volcano_ttest}}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=.85\textwidth]{sim_new/spike_in_statistic_comparison}
\caption{Comparing the summary statistics between methods, the proposed method performed the strongest. In terms of accuracy and specificity the three methods were close, with Limma and Anova showing slightly higher values. Accuracy and specificity were dominated by the large number of true negatives (background peptides) compared to the true positives (spike-in peptides). In terms of recall, the proposed approach out performed the other two methods, showing that it correctly labeled the most spike-in peptides. \label{fig:spike_stat}}
\end{figure}
